---
title: "SQL Project"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
  pdf_document: default
---

<div class="blame">
author: "Jonathan Lau"<br>
</div>


<hr>

<br>
<div class='emphasis'>


# Project Overview
*ACME Corporation - client database*<br>

## Project outline
• Basic analysis of stored data<br>
• Quantitative analysis of emotional state of corporation based on stored `feedbacks`.
• Practising coding best practice for R and SQL.

## R Packages used
• Keyring<br>
• RPostgreSQL<br>

## Description of data:
• PostgreSQL database<br>
• Contains three tables of information: details of `employees`, `teams` and `feedbacks`.<br>
• `Feedbacks` is text detailing employees' positive and negative experiences at work.
</div>
<br>

# Querying the SQL database

## Creating a local PostgreSQL database from `Terminal`

```{}
createdb acme_employees
psql -d acme_employees -f acme_employees.sql
```

## Initial look at database structure

Exploring database relationships from `Terminal` using `psql` command line interface


Three tables:
• `employees` (200 records)
• `feedbacks` (60 records)
• `teams` (11 records)

"Primary keys" (PK) and "Foreign keys" (FK)
• `employees`: PK id, FK `team_id` NOT NULL, referenced by `feedbacks`.
• `teams`: PK id, no FK, referenced by `employees`
• `feedbacks`: PK id, FK `employee_id` NOT NULL

Table relationships:
• One team has many employees
• One employee has many feedbacks


## Setting up database connection

Loading the required R libraries
```{r}
library(keyring)
library(RPostgreSQL)
```

Creating a local connection from R to acme_employees database
```{r}
# Using keychain on my Macbook

keyring_unlock(keyring = "local")

username <- key_get(service = "postgresql_username",
                    keyring = "local")

password <- key_get(service = "postgresql_password",
                    keyring = "local")

db_connect<- dbConnect(drv = PostgreSQL(max.con = 3),
                       user = username,
                       password = password,
                       dbname = "acme_employees",
                       host = "localhost")
```

Mustn't forget to remove my username and password or disconnect from the database at the end (!)
```{r}
rm(username, password)
```


## Pulling data

Creating a new data.frame containing details of all employees by querying SQL database
```{r}
#dbGetQuery
employees <- as.data.frame(
  dbGetQuery(db_connect,
             "SELECT * FROM employees")
  )
employees
```

## Adding new fields and joining tables

Retrieving records for all employees but this time adding a `full_name` field.
```{r}
dbGetQuery(db_connect,
"SELECT CONCAT(first_name, last_name) AS full_name,
* FROM employees"
)
```

Retrieving details of all employees and this time listing `team_name` instead of `team_id`

```{r}
#Using INNER JOIN on list of team names in relational database without creating aliases (for clarity)

dbGetQuery(db_connect,
"SELECT teams.name, employees.first_name, employees.last_name, employees.email 
FROM employees INNER JOIN teams
on employees.team_id = teams.id")
```

## Making queries

Breaking down number of employees per team.
```{r}
dbGetQuery(db_connect,
           "SELECT teams.name, COUNT(*)
           FROM teams LEFT JOIN employees
           on teams.id = employees.team_id
           GROUP BY teams.name
           ORDER BY COUNT(*) DESC")
```


AcmeCorp management wouldn't be able to email their employees easily if just using on this database as there are a lot of invalid email addresses there.

Identifying these employees,
```{r}
# looking for `email` fields that don't contain @ or are empty
dbGetQuery(db_connect,
           "SELECT first_name, last_name, email
           FROM employees
           WHERE email NOT LIKE '%@%' OR email IS NULL")
```

Finding employee that gave most recent feedback

```{r}
dbGetQuery(db_connect,
"SELECT feedbacks.date, employees.first_name, employees.last_name, feedbacks.message
FROM feedbacks LEFT JOIN employees ON feedbacks.employee_id = employees.id
   ORDER BY feedbacks.date DESC
   LIMIT 1
  ")

# Could also have used
# WHERE feedbacks.date = (SELECT MAX(date) FROM feedbacks)
```

One of the teams has a problem with employees whose first_name contains their first initial rather than their whole first name.

Finding name of the team which has the worst problem with this.
```{r}
dbGetQuery(db_connect,
"SELECT employees.first_name, employees.last_name, teams.name
FROM employees INNER JOIN teams
ON teams.id = employees.team_id
WHERE employees.first_name LIKE '_'")

dbGetQuery(db_connect,
"SELECT teams.name, count(teams.name)
FROM employees INNER JOIN teams
ON teams.id = employees.team_id
WHERE employees.first_name LIKE '_'
GROUP BY teams.name
ORDER BY count DESC NULLS LAST
LIMIT 1")
```

# Basic sentiment analysis by combining SQL and R coding


AcmeCorp would like to perform very basic ‘sentiment analysis’ on stored employee feedbacks. They have provided a set of ‘positive’ and ‘negative’ words to use to score feedback messages in this analysis.


## Scoring framework

• The score for each message should start at zero.
• Every time a ‘positive’ word occurs in a message, message’s score should increase by 1 point.
• Every time a ‘negative’ word occurs in a message, message’s score should decrease by 1 point.

List of positive words
```{r}
positive_words <- c("great", "good", "happy", "content", "fulfilled", "fulfilled", "positive", "positives", "cheerful", "nice", "ambition", "ambitions", "ambitious", "friendly", "progress", "joy", "fast", "achieve", "achieving", "achievements", "best", "better",  "unlimited", "pros", "competitive", "opportunity", "advancement", "friendly", "smart", "interesting", "challenging", "grow", "caring", "balance", "helpful")
```

List of negative words
```{r}
negative_words <- c("worse", "worsening", "worst", "slow", "sluggish", "impatient", "unfulfilled", "unfulfilling", "negative", "bad", "limited", "limit", "limitations", "cons", "frustrated", "frustrating", "frustration", "dull", "boring", "politics", "political", "ignore", "ignored", "ignoring", "sad", "depressed", "depressing", "nasty", "stress", "stresses", "stressed", "horrid", "horrible", "terrible", "deadline", "deadlines", "uncaring", "remote")
```


## Simplifying text structure for analysis

```{r}
# Querying database for feedback messages
feedback <- dbGetQuery(db_connect, "SELECT message FROM feedbacks")
feedback

# Extracting information
textwork <- feedback$message
textwork
length(textwork)

# setting everything to lowercase for ease of word recognition
textwork <- tolower(textwork)

# Using gsub to flag and remove punction
clean_text <- gsub("[[:punct:]]"," ", textwork)

```


## Text analysis


Scoring the messages
```{r}
# Creating a function to programatically score the messages

score_message <- function(message, positive_words, negative_words){
  message_words <- unlist(strsplit(message, split = " ", fixed = TRUE))
  message_score <- 0
  
  for (word in positive_words){
    message_score <- message_score + sum(message_words == word)
  }
  
  for (word in negative_words){
    message_score <- message_score - sum(message_words == word)
  }
  
  return(message_score)
}
```

Cleaning and then adding the scores to the original feedback messages
```{r}
score <- numeric()
for (message in feedback[, 'message']){
  message <- gsub("[[:punct:]]"," ", message)
  message_score <- score_message(message, positive_words, negative_words)
  score <- c(score, message_score)
}

# add the scores 
feedback['score'] <- score
feedback
```


Finding the most positive and negative employees


```{r}
most_positive <- feedback[, "score"] == max(feedback[, 'score'])

mostpos_employees_teams <- feedback[most_positive, "team_name"]
mostpos_employees_teams
```

Finding the most positive teams

```{r}

```

## Don't forget to disconnect from the database!

```{r}
#dbDisconnect(db_connect)
```



```{r}
#It's good practice to code out your database discconection first as a reminder.  Don't want to upset your colleagues by monopolising a connection if you're not using it.  Worth commenting it out to avoid disconnecting by accident though.

dbDisconnect(conn = db_connect)
```

